\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-09-15 analysis/paper/paper.Rmd
update CS3

paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
```

@@

In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves. <!--In the next two case studies, we explore more advanced features.--> For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function. To do this, we simply replace logistic regression with a penalized counterpart. In \pkg{casebase}, we achieve this by using the \pkg{glmnet} package [@friedman2010jss].

To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support]. The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill. The original data is available online from the Department of Biostatistics at Vanderbilt University [@harrell_2020]. The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package. For more information about this dataset, the reader is encouraged to look at the documentation in our package. A description of the variables can be found in Table \ref{tab:support1}, and a breakdown of each categorical variable appears in Table \ref{tab:support2}.
\begin{tabular}{ccccc}
\hline
\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
\hline
age           & Age                                      &                 & double           & 0            \\
death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
sex           &                                          & 2               & integer          & 0            \\
slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
d.time        & Days of Follow-Up                        &                 & double           & 0            \\
dzgroup       &                                          & 8               & integer          & 0            \\
dzclass       &                                          & 4               & integer          & 0            \\
num.co        & number of comorbidities                  &                 & double           & 0            \\
edu           & Years of Education                       &                 & double           & 202          \\
income        &                                          & 4               & integer          & 349          \\
scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
race          &                                          & 5               & integer          & 5            \\
sps 	        & support physiology score day 3           &	               & double           &	1            \\
aps           & APS III no coma, imp bun,uout for ph1,D3 &	               & double	          & 1            \\
hday	        & Day in Hospital at Study Admit	         &                 & double           & 0            \\	
diabetes      &	Diabetes (Com 27-28, Dx 73)	             &                 & double	          & 0            \\
dementia      & Dementia (Comorbidity 6)                 &	               & double	          & 0            \\
ca		        & Cancer State                             & 3               & integer	        & 0            \\
meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
bili          & Bilirubin Day 3                          &                 & double           & 297          \\
crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
sod           & Serum sodium Day 3                       &                 & double           & 0            \\
ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
glucose       & Glucose Day 3                            &                 & double           & 470          \\
bun           & BUN Day 3                                &                 & double           & 455          \\
urine         & Urine Output Day 3                       &                 & double           & 517          \\
adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\
adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
\end{tabular}
\caption{Description of each variable in the SUPPORT dataset.}
\label{tab:support1}

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
ca       & Yes                                           \\
         & No                                            \\
         & Metastatic                                    \\
\end{tabular}
\caption{Description of each level within each categorical variable in the SUPPORT dataset.}
\label{tab:support2}
\end{table}

<!--Two variables are SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology score (aps). These scores were developed for a period of 180 days, which makes our comparison over 5.56 years exploratory. Therefore, we will assess these two models before using either as our baseline, by comparing them to a full model using all variables. Also, our single imputation introduces bias in the data. As the goal here is purely demonstrative, this bias will be consistent with any of the following models. As such, their comparison is still valid, though any interpretation requires an awareness of said bias. The SUPPORT dataset will be used to demonstrate regularization within casebase, while comparing their absolute risk predictions for a new individual.-->

In the original study, the authors developed two scores for predicting the outcome. These scores are available as part of the data: SUPPORT day 3 physiology score (\code{sps}) and APACHE III day 3 physiology score (\code{aps}). Using these scores, we will explore three Cox regression scenarios as a baseline comparison for \pkg{casebase}: \code{sps} only, \code{aps} only, and all covariates excluding \code{sps} and \code{aps}. We will compare the three models using a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass]. The Brier score can be used to assess both discrimination and calibration in a single model. Moreover, it provides a metric that can be used to compare parametric and semi-parametric survival models. We will use 95% of the observations to train our models, and the remaining 5% to compute absolute risk curves and Brier scores.

@@

In Figure \ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves. On the other hand, Figure \ref{fig:brier1} (B) shows that the full model has the lowest Brier score overall. For this reason, we will use this model in our comparison between regularized Cox regression, Kaplan-Meier estimation, and the approach from \pkg{casebase}.

<!--Before generating an absolute risk curve using regularized estimation, we will explore three options as a baseline comparison. The original study compared these scores over a span of 180 days, which may not be comparable over the full 5.56 years. Therefore, we will examine three Cox regression scenarios: sps only, aps only and all covariates excluding sps and aps. Next, we will compare each modeling using the Brier score [@graf1999ass]. The Brier score can demonstrate both discrimination and calibration in a single model [@graf1999ass] and provide a comparable metric between semi-parametric and parametric methods in the survival context. For all the following analyses in this case study, we will use 80% of the observations to train our models, while 20% will be used to generate absolute risk curves, all of which will be averaged to demonstrate each model's predictive ability. Over 5.56 years, \ref{fig:baselines} demonstrates that all the cox models have similar, if not identical average predictions. Figure \ref{fig:brier1} examines the brier scores as they change over time. As the full model has the lowest brier score on overall, it will be used in the following absolute risk comparison to regularized cox, regularized estimation and the unadjusted Kaplan-Meier curve.-->

```{r baselinePlotsPrep, eval=eval_cs3, echo = FALSE}
# Define colour palette for this section
q4 <- qualitative_hcl(4, palette = "Dark 3")

# 1. Cox with only sps
coxSPS <- survival::coxph(Surv(time = d.time, event = death) ~ sps, 
                          data = train, x = TRUE)
abcoxSPS <- survival::survfit(coxSPS, newdata = test)
# 2. Cox with only aps
coxAPS <- survival::coxph(Surv(time = d.time, event = death) ~ aps, 
                          data = train, x = TRUE)
abcoxAPS <- survival::survfit(coxAPS, newdata = test)
# 3. Cox with everything but sps and aps
coxFull <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps, 
                           data = train, x = TRUE)
abcoxFull <- survival::survfit(coxFull, newdata = test)
# 4. Kaplan-Meier
KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test,x=TRUE)
abKM <- survival::survfit(KM)
abKMcompare <- data.frame(
  Time = abKM$time,
  KM = 1 - abKM$surv
)

# Combine all CI estimates from Cox models
baseCoxComparisonData <- data.frame(
  Time = abcoxSPS$time,
  SPS = 1 - rowMeans(abcoxSPS$surv),
  APS = 1 - rowMeans(abcoxAPS$surv),
  Full = 1 - rowMeans(abcoxFull$surv)
)
```

```{r baselines, echo=FALSE, eval=eval_cs3}
cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) +
  geom_step(aes(y = APS, colour = "APS")) +
  geom_step(aes(y = Full, col = "Full")) +
  geom_step(aes(y = SPS, colour = "SPS")) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
  paper_gg_theme + 
  scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) + 
  scale_y_continuous(labels = scaleFUNy, n.breaks = 8)
```

```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve. (A) Cumulative Incidence curves. For a given model, the plotted curve is the average curve of each individual in the study. Each follow-up time seen in the study was used. There are minimal differences between the curves. APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better. The full model performs best overall."}
# Extract time points
times <- sort(unique(test$d.time))
# Compute Brier scores
brierAPS <- Score(list("PredictionModel" = coxAPS), data = test, 
                  formula = Hist(d.time, death != 0) ~ 1, summary = "ipa", 
                  se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
brierSPS <- Score(list("PredictionModel" = coxSPS), data = test, 
                  formula = Hist(d.time, death != 0) ~ 1, summary = "ipa", 
                  se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)

# First fix NA coefficients, then compute Brier score
coxFull$coefficients[which(is.na(coxFull$coefficients))] <- 0
brierFull <- Score(list("PredictionModel" = coxFull), data = test, 
                   formula = Surv(d.time, death != 0) ~ 1, summary = "ipa", 
                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
brierNull <- Score(list("PredictionModel" = KM), data = test, 
                   formula = Surv(d.time, death != 0) ~ 1, summary = "ipa", 
                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)

# Combine all results for plotting
brierPlotter <- rbind(
  data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"], 
             times = times, model = "APS"),
  data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"], 
             times = times, model = "Full"),
  data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"], 
             times = times, model = "SPS"),
    data.frame(brierScore = brierNull$Brier$score$Brier[brierNull$Brier$score$model == "PredictionModel"], 
             times = times, model = "K-M")
)

cox_brier <- ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) +
  geom_line() +
  xlab("Follow-up time (Years)") +
  ylab("Brier score") +
  labs(color = "Models") +
  # ggtitle("Brier scores over time") +
  scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) + 
  paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1)))+scale_y_continuous(labels=scaleFUNy,n.breaks=8)

# Combine both plots using cowplot
plot_row <- cowplot::plot_grid(cox_ci + theme(legend.position = 'none'),
                               cox_brier + theme(legend.position = 'none'),
                               labels = c("A", "B"))
legend <- cowplot::get_legend(
  # create some space to the left of the legend
  cox_ci + theme(legend.box.margin = margin(0, 0, 0, 12))
)

cowplot::plot_grid(plot_row, legend, rel_heights = c(3, .4), nrow = 2)
```

<!--To get the absolute risk following regularized cox regression, we had to work around limitations in existing packages. To make use of the \pkg{survival} package, we first estimate the hazard coefficients using thecoxnet implementation for lasso penalization in \pkg{glmnet} [@regpathcox]. Next, we fit a cox model using the survival package with the selected variables only, from our regularized fit. Finally, we replaced the coefficients in the survival object with those from our glmnet fit, allowing us to make use of the absolute risk tools available in the \pkg{survival} package.-->

For our penalized model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution. For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}. In order to use this interface, we first create a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all remaining covariates we would like to include in the model space. We used lasso penalization by setting \code{alpha = 1}. Moreover, we use the argument \code{penalty.factor} to force the time variable into the final model.

```{r supportCB_fit, eval=eval_cs3, echo=TRUE, cache=FALSE}
x <- model.matrix(death ~ . - d.time - aps - sps, 
                  data = train)[, -c(1)] # Remove intercept
y <- data.matrix(subset(train, select = c(d.time, death)))

# Regularized logistic regression to estimate hazard
cbFull <- casebase::fitSmoothHazard.fit(x, y,
  family = "glmnet",
  time = "d.time", event = "death",
  formula_time = ~ log(d.time), alpha = 1,
  ratio = 10, standardize = TRUE,
  penalty.factor = c(0, rep(1, ncol(x)))
)
```

Next, we can use the object \code{cbFull} to estimate the absolute risk curve on the test data. We first create a matrix \code{newx} containing the relevant variables. Next, using \code{times}, we specify the time points at which the cumulative incidence will be computed. Finally, using the parameter \code{s = "lambda.1se"}, we specify at which value of the tuning parameter we want to retrieve the coefficient estimates.

```{r supportCB_abs, eval=eval_cs3, echo=TRUE, cache=TRUE}
# Estimating the absolute risk curve using the newdata parameter
newx <- model.matrix(death ~ . - d.time - aps - sps, 
                  data = test)[, -c(1)]
times <- sort(unique(test$d.time))
#times<-seq(from=min(times),to=max(times),length.out = 100)

abCbFull <- casebase::absoluteRisk(cbFull,
  time = times,
  newdata = newx,
  s = "lambda.1se",
  method = "numerical"
)
```

@@

# The coefficients of this object will be replaced with the estimates from the
# original coxNet. Doing so makes it so that everything is invalid aside from
# the coefficients. In this case, all we need to estimate the absolute risk is
# the coefficients. Std. error would be incorrect here, if we were to draw error
# bars.
coxNet_coefnames <- names(coxNet$coefficients)
coxNet$coefficients <- nonzero_coef_cox@x
names(coxNet$coefficients) <- coxNet_coefnames

# Fitting absolute risk curve for cox+glmnet
abcoxNet <- survival::survfit(coxNet, type = "breslow", 
                              newdata = as.data.frame(newx[, nonzero_covariate_cox$X1]))

# Combine CI estimates from both penalized methods
abCN <- data.frame(
  Time = abcoxNet$time,
  absRisk = 1 - rowMeans(abcoxNet$surv)
)
abCN$Model <- "Penalized Cox"

abCB <- data.frame(Time = abCbFull[, 1],
                   absRisk = rowMeans(abCbFull[, -c(1)]))
abCB$Model <- "Penalized CB"

ab <- rbind(abCB, abCN)

Our first comparison plot examines the covariates that were selected between all three hazard functions. Note that Cox regression without penalization selected more covariates than the other two, but only the intersection of coefficients with values greater than 0 are displayed in \ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have larger estimated coefficients. This is expected, as there are no restrictions on the size of coefficients. The smallest coefficient estimates are found in penalized case-base model, aside from cancer state (cano).

```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models. If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity. Shrinkage occurs due to penalization."}
library(dotwhisker)
library(broom)
library(dplyr)

@@

estimatescb <- data.table::setDT(as.data.frame(coef(cbFull)[-c(1, 2), 1]), 
                     keep.rownames = TRUE)
# estimatescb$std.error<-0
estimatescb$Model <- "Pen. CB"
colnames(estimatescb) <- c("term", "estimate", "model")

estimatescoxFull <- data.table::setDT(as.data.frame(coef(coxFull)), 
                          keep.rownames = TRUE)
# estimatescoxFull$std.error<-0
estimatescoxFull$Model <- "Cox"
colnames(estimatescoxFull) <- c("term", "estimate", "model")

# Clean up the data and the variable names
lolliplotDots <- rbind(estimatescoxFull, estimatescoxNet, estimatescb)
lolliplotDots$conf.low <- lolliplotDots$estimate
lolliplotDots$conf.high <- lolliplotDots$estimate
lolliplotDots$conf.high[lolliplotDots$estimate < 0] <- 0
lolliplotDots$conf.low[lolliplotDots$estimate > 0] <- 0
lolliplotDots$term <- gsub("\`", "", as.character(lolliplotDots$term))

@@

lolliplotDots <- lolliplotDots[-which(is.na(lolliplotDots$estimate)), ]
dwplot(lolliplotDots) +
  theme(
    axis.text.y = element_text(size = 11, angle = 0, hjust = 1, vjust = 0),
    strip.text.x = element_blank(),
    strip.background = element_rect(colour = "white", fill = "white"),
    legend.position = c(.855, 0.15), legend.text = element_text(size = 11)
  ) +
  paper_gg_theme +  
  # ggtitle("Coefficient plot") +
  labs(color = "Models") +
  scale_color_manual(values = q4[1:3])
```

These values do not inform us of model performance. To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time. The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A). There is separation between the curves, but it is minimal. To better demonstrate the differences in performance between the models, we compare Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized case-base model compared to both Cox regression models. The differences, however, are minimal. All three outperform the Kaplan-Meier estimator.

```{r cs3Final, echo=FALSE, eval=eval_cs3}
reg_ci <- ggplot(ab, aes(Time)) +
  geom_step(data = baseCoxComparisonData, mapping = aes(x = Time,y = Full, colour = "Cox")) +
  geom_line(aes(y = absRisk, colour = Model)) +
  geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) +
  labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
  # ggtitle("Absolute risk curves") +
  scale_color_manual(values = q4, breaks = c("K-M", "Cox", "Penalized Cox", "Penalized CB")) + 
  paper_gg_theme + 
  scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) + 
  scale_y_continuous(labels = scaleFUNy, n.breaks = 9)
```

@@

```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier. (A)  Probabilities of death as it changes over follow-up time. (B) Brier scores as they change over time, where a lower score corresponds to better performace. Penalized case-base estimation produced the lowest results over follow-up time.", warning = FALSE}
# Compute Brier score for penalized models
brierFinalResults <- Score(list("Penalized Cox" = coxNet, "Penalized CB" = cbFull),
                           data = cbind(subset(test, select = c(d.time, death)),
                                        as.data.frame(newx)), 
                           formula = Hist(d.time, death != 0) ~ 1, summary = NULL, 
                           se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times)
#brierFinalResults$models<-as.character(brierFinalResults$models)
# Combine Brier scores between unpenalized and penalized models before plotting

data_brier <- bind_rows(
  brierFinalResults$Brier$score,
  filter(brierFull$Brier$score, model == "PredictionModel")
  ) %>% 
  mutate(model = as.character(model), # First return to string, so types match
         model = case_when(
    model == "Null model" ~ "K-M",
    model == "PredictionModel" ~ "Cox",
    TRUE ~ model
  ),
  model = factor(model, levels = c("Cox", "Penalized Cox", "Penalized CB", "K-M"),
                 labels = c("Cox", "Pen. Cox", "Pen. CB", "K-M")))

reg_brier <- ggplot(data = data_brier, mapping = aes(x = times, y = Brier, col = model)) +
  geom_line() +
  xlab("Follow-up time (years)") +
  ylab("Brier score") +
  # ggtitle("Brier score over time") +
  labs(color = "Models") +
  paper_gg_theme +  
  scale_color_manual(values = q4) + 
  scale_x_continuous(labels = scaleFUNx, 
                     breaks = round(seq(0, 5, by = 1))) + 
  scale_y_continuous(labels = scaleFUNy, n.breaks = 8)

@@

These values do not inform us of model performance. To do so, we take our hazard functions and use them to predict the absolute risk of everyone in our test set, over time. The probabilities over time for each individual are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A). There is separation between the curves, but it is minimal. To better demonstrate the differences in performance between the models, we compare Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our penalized logistic regression model compared to both Cox regression models. The differences, however, are minimal. All three outperform the Kaplan-Meier model.

@@

In the previous case studies, we only considered covariates that were fixed at baseline. In this next case study, we will use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-dependent exposure. As an example that already appeared in the literature, case-base sampling was used to study vaccination safety, where the exposure period was defined as the week following vaccination [@saarela2015case]. Hence, the main covariate of interest, i.e. exposure to the vaccine, was changing over time. In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.

@@

As before, we can look at the population-time plot for a graphical summary of the event incidence. Here, we coloured the exposed person-time (i.e. after transplant) in a darker shade of gray. As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.

@@

```{r stanford_cb, eval = eval_cs4}
cb_data <- sampleCaseBase(jasa,
  time = "futime",
  event = "fustat", ratio = 100
)
```

@@

>>>>

paper_gg_theme <- theme_bw(base_size = 10) + theme(legend.position = "bottom")
# Define colour palette for CS 3
q4 <- qualitative_hcl(4, palette = "Dark 3")
names(q4) <- c("Cox", "Pen. Cox", "Pen. CB", "K-M")
```

@@

In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves. For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function. To do this, we simply replace logistic regression with a penalized counterpart. In \pkg{casebase}, we achieve this by using the \pkg{glmnet} package [@friedman2010jss].

To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support]. The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill. The original data is available online from the Department of Biostatistics at Vanderbilt University [@harrell_2020]. The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package. For more information about this dataset, the reader is encouraged to look at the documentation in our package.

In this section, we compare our penalized case-base approach to two other models: Cox regression, and penalized Cox regression (using the \pkg{glmnet} package). In all three models, we use all covariates except \code{sps} and \code{aps}. These two variables correspond to scores for predicting the outcome that were developed as part of the original study.

To compare the performance of our models, we first divide the data into a training and a test data, with 95% of the observations used for training all three models. To assess both discrimination and calibration, we use a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass]. The Brier score can be used with both parametric and semi-parametric models. We use the \pkg{riskRegression} to compute these scores for all models.

@@

```{r supportCox_fit, eval=eval_cs3, echo=FALSE, cache=FALSE}
# Cox with everything but sps and aps
cox <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps,
                           data = train, x = TRUE)
```

For our penalized case-base model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution. For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}. In order to use this interface, we first create a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all remaining covariates we would like to include in the model space. We used lasso penalization by setting \code{alpha = 1}. Moreover, we use the argument \code{penalty.factor} to give a regularization weight of 0 to the time variable, and therefore force it into the final model.

```{r supportCB_fit, eval=eval_cs3, echo=TRUE, cache=FALSE}
# Create matrices for inputs
x <- model.matrix(death ~ . - d.time - aps - sps, 
                  data = train)[, -c(1)] # Remove intercept
y <- data.matrix(subset(train, select = c(d.time, death)))

# Regularized logistic regression to estimate hazard
pen_cb <- casebase::fitSmoothHazard.fit(x, y,
  family = "glmnet",
  time = "d.time", event = "death",
  formula_time = ~ log(d.time), alpha = 1,
  ratio = 10, standardize = TRUE,
  penalty.factor = c(0, rep(1, ncol(x)))
)
```

<!-- Next, we can use the object \code{cbFull} to estimate the absolute risk curve on the test data. We first create a matrix \code{newx} containing the relevant variables. Next, using \code{times}, we specify the time points at which the cumulative incidence will be computed. Finally, using the parameter \code{s = "lambda.1se"}, we specify at which value of the tuning parameter we want to retrieve the coefficient estimates. -->

@@

# The coefficients of this object will be replaced with the estimates from the
# original coxNet. Doing so makes it so that everything is invalid aside from
# the coefficients. In this case, all we need to estimate the absolute risk is
# the coefficients. Std. error would be incorrect here, if we were to draw error
# bars.
coxNet_coefnames <- names(coxNet$coefficients)
coxNet$coefficients <- nonzero_coef_cox@x
names(coxNet$coefficients) <- coxNet_coefnames
```

Our first comparison plot examines the covariates that were selected between all three hazard functions. In \ref{fig:cs3lolliPlot}, we show the coefficient estimates for covariates that we selected by both penalized Cox and penalized case-base. From this plot, we can clearly see the shrinkage effect of using penalized estimation.

```{r coefplots, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models. Only covariates common among all three models were retained. Shrinkage occurs due to penalization."}
library(dotwhisker)
library(broom)
library(dplyr)

@@

estimatescb <- data.table::setDT(as.data.frame(coef(pen_cb)[-c(1, 2), 1]),
                                 keep.rownames = TRUE)
# estimatescb$std.error<-0
estimatescb$Model <- "Pen. CB"
colnames(estimatescb) <- c("term", "estimate", "model")

estimatescox <- data.table::setDT(as.data.frame(coef(cox)), 
                                  keep.rownames = TRUE)
# estimatescox$std.error<-0
cox$coefficients[which(is.na(cox$coefficients))] <- 0
estimatescox$Model <- "Cox"
colnames(estimatescox) <- c("term", "estimate", "model")

# Clean up the data and the variable names
lolliplotDots <- rbind(estimatescox, estimatescoxNet, estimatescb)
lolliplotDots$conf.low <- lolliplotDots$estimate
lolliplotDots$conf.high <- lolliplotDots$estimate
lolliplotDots$conf.high[lolliplotDots$estimate < 0] <- 0
lolliplotDots$conf.low[lolliplotDots$estimate > 0] <- 0
lolliplotDots$term <- gsub("\`", "", as.character(lolliplotDots$term))

@@

lolliplotDots <- lolliplotDots[-which(is.na(lolliplotDots$estimate)), ]
dwplot(lolliplotDots) +
  theme(
    axis.text.y = element_text(size = 11, angle = 0, hjust = 1, vjust = 0),
    strip.text.x = element_blank(),
    strip.background = element_rect(colour = "white", fill = "white"),
    legend.position = c(.855, 0.15), legend.text = element_text(size = 11)
  ) +
  paper_gg_theme +  
  # ggtitle("Coefficient plot") +
  labs(color = "Models") +
  scale_color_manual(values = q4)
```

Next, we compare the cumulative incidence estimation over the test set. The probabilities over time for each observation are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A). We can see some minimal differences between the three models, with the Kaplan-Meier estimate giving a lower estimate than our three models. We also note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. As described above, we compare the performance between the models by computing the Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models. On the other hand, the differences between these three models are minimal.

```{r support_abs, eval=eval_cs3, echo=FALSE}
# Absolute Risks
newx <- model.matrix(death ~ . - d.time - aps - sps,
                     data = test)[, -c(1)]
times <- sort(unique(test$d.time))
# 1. Unpenalized Cox
abcox <- survival::survfit(cox, newdata = test)

# 2. Penalized Cox
abcoxNet <- survival::survfit(coxNet, type = "breslow", 
                              newdata = as.data.frame(newx[, nonzero_covariate_cox$X1]))

# 3. Penalized Case-base
abpen_cb <- casebase::absoluteRisk(pen_cb,
  time = times,
  newdata = newx,
  s = "lambda.1se",
  method = "numerical"
)

# 4. Kaplan-Meier
KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test,
                      x = TRUE)
abKM <- survival::survfit(KM)

# Combine absolute risk estimates
data_absRisk <- bind_rows(
  data.frame(Time = abcox$time, Prob = 1 - rowMeans(abcox$surv),
             Model = "Cox"),
  data.frame(Time = abcoxNet$time, Prob = 1 - rowMeans(abcoxNet$surv),
             Model = "Pen. Cox"),
  data.frame(Time = abpen_cb[, 1], Prob = rowMeans(abpen_cb[, -c(1)]),
             Model = "Pen. CB"),
  data.frame(Time = abKM$time, Prob = 1 - abKM$surv,
             Model = "K-M")
) %>% 
  mutate(Model = factor(Model, levels = names(q4)))
```

```{r absRiskPlot, echo=FALSE, eval=eval_cs3}
reg_ci <- ggplot(data_absRisk, aes(Time, Prob)) +
  geom_line(aes(colour = Model)) +
  labs(y = "Probability of death", x = "Follow-up time (years)", color = "Models") +
  scale_color_manual(values = q4) + 
  expand_limits(y = 1) +
  paper_gg_theme + 
  scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) + 
  scale_y_continuous(labels = scaleFUNy, n.breaks = 9)
```

@@

```{r support_Brier, eval=eval_cs3, echo=FALSE}
# Brier score
# 1. Unpenalized models
# First fix NA coefficients, then compute Brier score
brierCoxKM <- Score(list("Cox" = cox,
                         "K-M" = KM), data = test,
                   formula = Hist(d.time, death != 0) ~ 1, summary = NULL,
                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, 
                   times = times)

# 2. Penalized models
brierPenalized <- Score(list("Pen. Cox" = coxNet, 
                             "Pen. CB" = pen_cb),
                        data = cbind(subset(test, select = c(d.time, death)),
                                     as.data.frame(newx)),
                        formula = Hist(d.time, death != 0) ~ 1, summary = NULL, 
                        se.fit = FALSE, metrics = "brier", contrasts = FALSE, 
                        times = times)
```

```{r riskregressionBrier, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier. (A)  Probabilities of death as it changes over follow-up time. (B) Brier scores as they change over time, where a lower score corresponds to better performace.", warning = FALSE}
# Combine scores
data_brier <- bind_rows(
  brierCoxKM$Brier$score %>% 
    mutate(model = as.character(model)) %>% 
    filter(model != "Null model"),
  brierPenalized$Brier$score %>% 
    mutate(model = as.character(model)) %>% 
    filter(model != "Null model")
  ) %>% 
  mutate(model = factor(model, levels = c("Cox", "Pen. Cox", "Pen. CB", "K-M")))

reg_brier <- ggplot(data = data_brier, aes(x = times, y = Brier, col = model)) +
  geom_line() +
  xlab("Follow-up time (years)") +
  ylab("Brier score") +
  labs(color = "Models") +
  paper_gg_theme +  
  scale_color_manual(values = q4) + 
  scale_x_continuous(labels = scaleFUNx, 
                     breaks = round(seq(0, 5, by = 1))) + 
  scale_y_continuous(labels = scaleFUNy, n.breaks = 8)

@@

In brief, case-base sampling can be used in conjunction with penalized logistic regression to perform variable selection with survival models.

@@

In the previous case studies, we only considered covariates that were fixed at baseline. In this next case study, we use the Stanford Heart Transplant data [@clark1971cardiac,@crowley1977covariance] to show how case-base sampling can also be used in the context of time-dependent exposure. This feature of case-base sampling has been explored in the literature, in the context of vaccination safety  [@saarela2015case]. In this study, the exposure period was defined as the week following vaccination. Hence, the main covariate of interest, i.e. exposure to the vaccine, was changing over time. In this context, case-base sampling offers an efficient alternative to nested case-control designs or self-matching.

@@

As above, we can look at the population-time plot for a graphical summary of the event incidence. Here, we colour the exposed person-time (i.e. after transplant) in a darker shade of gray. As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.

@@

```{r stanford_cb, eval = eval_cs4, echo = TRUE}
cb_data <- sampleCaseBase(jasa,
  time = "futime",
  event = "fustat", ratio = 100
)
```

@@

<!-- # Appendix -->
